Bayesian Restricted Likelihood Methods: Conditioning on Insufficient Statistics in Bayesian Regression

### 2018-08-19 revision_1/paper.tex
finishing up hiearchical simulation example

\section{Simulated Data}
For this study, the simulated data comes from the following hierarchical data generating model:
\begin{align}
\label{gensim2}
\begin{split}
& \theta_{i}  \sim   N(\mu, \tau^{2}),  \ i = 1, 2, \dots, 90  \\ 
& y_{ij} \sim (1-p_{i})N(\theta_{i}, \sigma^{2}) + p_{i}N(\theta_{i}, m_{i}\sigma^{2}),\  j = 1, 2,..., n_{i}
\end{split}
\end{align}
with $\mu = 0, \tau^{2} = 1, \sigma^{2} = 4$. The values of $p_{i}, m_{i}$, and $n_{i}$ depend on the group and are formed using 5 replicates of the full factorial design over factors $p_{i},m_{i},n_{i}$ with levels $p_{i} = .1, .2, .3$, $m_{i} = 9, 25$, and $n_{i} = 25, 50, 100$. Thus, the 90 groups have a varying level of outlier contamination and sample size. We wish to build models that offer good prediction for the good portion of data within each group. The full model for fitting is a corresponding normal model without contamination:
\begin{equation}
\label{fullsim2}
\begin{split}
& \mu \propto 1, \  \tau^{2} \propto \tau^{-2}, \\
& \theta_{i}\sim N(\mu, \tau^{2}), \  \sigma^{2}_{i} \sim IG(a_{s}, b_{s}),  \ i = 1, 2, \dots, 90, \\ 
& y_{ij}\sim  N(\theta_{i},\sigma^{2}_{i}), \ j = 1, 2, \dots, n_{i}
\end{split}
\end{equation}
For the restricted likelihood versions we condition on robust M-estimators of location and scale in each group: $T_{i}(\by_{i}) = (\hat\theta_{i}, \hat\sigma^{2}_{i}), i = 1, 2, ..., 90$. The two versions use Huber's and Tukey's $\psi$ function, while both versions use Huber's $\chi$ function in $\eqref{Mest}$.

To complete the specification of model \eqref{fullsim2} we must specify $a_{s}$ and $b_{s}$. For this we use a variety of values representing different levels of prior knowledge. For each we set $b_{s} = 4a_{s}c_{s}$ resulting in a prior mean for each $\sigma^{2}_{i}$ of $\frac{4c_{s}a_{s}}{a_{s}-1}, \ a_{s} >1$. The larger $a_{s}$, the more informative the prior. With $c_{s} = 1$ the shrinkage (for large $a_{s}$) is to the true value of $\sigma^{2} = 4$ of the variance of the good component of data in the data generating model. We consider $a_{s} = 1.25,  5, 10,$ and $c_{s} = 1/2, 1, 2$. Values of $c_{s} \neq 1$ will result in shrinkage towards the wrong value of $\sigma^{2}$.

@@

The models are compared using Kullback-Leibler (KL) divergence from the distribution of good data to the posterior predictive distribution in order to assess the predictive capability of the models for the good part of the data. Specifically, the $i^{th}$ group of the $k^{th}$ simulated data set $\by_{k}$ compute:
\begin{equation}
\label{kl}
KL^{(M)}_{ik} = \int \log \frac{f(\tilde y | \theta_{i}, \sigma^{2})}{f_{i}(\tilde y | M, \by_{k})}  f(\tilde y | \theta_{i}, \sigma^{2}) \ dy
\end{equation}
where $M$ indexes the fitting model, $f(\tilde y | \theta_{i}, \sigma^{2})$ and is the pdf of a $N(\theta_{i}, \sigma^{2})$ at $\tilde y$ with $\sigma^{2} = 4$. For the Bayesian models ${f_{i}(\tilde y | M, \by_{k})} = \int f(\tilde y |\theta_{i}, \sigma_{i}^{2}) \pi(\theta_{i}, \sigma^{2}_{i} | M, \by_{k})$ where $ f(\tilde y |\theta_{i}, \sigma_{i}^{2})$ is again a normal pdf and  $\pi(\theta_{i}, \sigma^{2}_{i} | M, \by_{k})$ is the posterior for the $i^{th}$ group model parameters under model $M$ for the $k^{th}$ data set. $M$ denotes either the full normal theory model \eqref{fullsim2} or one of the two restricted likelihood versions discussed above, where conditioning is only on the summary statistics (not the complete $\by_{k}$). For the classical robust fits, we set $f_{i}(\tilde y | M, \by_{k})$ a $N(\hat\theta_{i}, \hat\sigma^{2}_{i})$ at  $\tilde y$ as a plug in estimator for the predictive distribution. The overall mean $\overline{KL}^{(M)}_{{\cdot}{\cdot}} = \frac{1}{90K} \sum_{k = 1}^{K} \sum_{i=1}^{90} KL^{(M)}_{ik}$ is used to compare the models with $SE(\overline{KL}^{(M)}_{{\cdot} k}) = \sqrt{\frac{K}{K-1}\sum_{k = 1}^{K} (\overline{KL}^{(M)}_{{\cdot} k} - \overline{KL}^{(M)}_{{\cdot}{\cdot}})^{2}}$ used to summarize sampling variation.

Figure \ref{} displays the results for each $a_{s} = 1.25,  5, 10,$ and $c_{s} = 1/2, 1, 2$ over $K = ?? $ data sets.

\begin{figure}[H]
\centering
\includegraphics[width = 6in, height = 4in]{kl_sim2_facet_scale.png}
\end{figure}
%
%\begin{figure}[H]
%\centering
%\includegraphics[scale = .3]{kl_sim2_facet_as.png}
%\end{figure}

>>>>

\section{Simulated Data}
We study the performance of the restricted likelihood in a hierarchical setting contaminated with outliers. Specifically, simulated data come from the following data generating model:
\begin{align}
\label{gensim2}
\begin{split}
& \theta_{i}  \sim   N(\mu, \tau^{2}),  \ i = 1, 2, \dots, 90  \\ 
& y_{ij} \sim (1-p_{i})N(\theta_{i}, \sigma^{2}) + p_{i}N(\theta_{i}, m_{i}\sigma^{2}),\  j = 1, 2,..., n_{i}
\end{split}
\end{align}
with $\mu = 0, \tau^{2} = 1, \sigma^{2} = 4$. The values of $p_{i}, m_{i}$, and $n_{i}$ depend on the group and are formed using 5 replicates of the full factorial design over factors $p_{i},m_{i},n_{i}$ with levels $p_{i} = .1, .2, .3$, $m_{i} = 9, 25$, and $n_{i} = 25, 50, 100$. This results in 90 groups that have varying levels of outlier contamination and sample size. We wish to build models that offer good prediction for the good portion of data within each group. The full model for fitting is a corresponding normal model without contamination:
\begin{equation}
\label{fullsim2}
\begin{split}
& \mu \propto 1, \  \tau^{2} \propto \tau^{-2}, \\
& \theta_{i}\sim N(\mu, \tau^{2}), \  \sigma^{2}_{i} \sim IG(a_{s}, b_{s}),  \ i = 1, 2, \dots, 90, \\ 
& y_{ij}\sim  N(\theta_{i},\sigma^{2}_{i}), \ j = 1, 2, \dots, n_{i}.
\end{split}
\end{equation}
For the restricted likelihood versions we condition on robust M-estimators of location and scale in each group: $T_{i}(\by_{i}) = (\hat\theta_{i}, \hat\sigma^{2}_{i}), i = 1, 2, ..., 90$. The two versions use Huber's and Tukey's $\psi$ function, while both versions use Huber's $\chi$ function in $\eqref{Mest}$. These estimators are implemented in the R function \texttt{MASS::rlm} and \ref{huber2009} provides further details on them.

To complete the specification of model \eqref{fullsim2} we must specify $a_{s}$ and $b_{s}$. For this we use a variety of values representing different levels of prior knowledge. For each we set $b_{s} = 4a_{s}c_{s}$ resulting in a prior mean for each $\sigma^{2}_{i}$ of $\frac{4c_{s}a_{s}}{a_{s}-1}, \ a_{s} >1$. The precision is $\frac{(a_{s} -1)^{2}(a_{s}-2)}{(4c_{s}a_{s})^{2}}$; meaning the larger $a_{s}$, the more informative the prior. With $c_{s} = 1$ the shrinkage (for large $a_{s}$) is to the true value of $\sigma^{2} = 4$ of the variance of the good component of data in the data generating model. We consider $a_{s} = 1.25,  5, 10$ and $c_{s} = 1/2, 1, 2$. %Values of $c_{s} \neq 1$ will result in shrinkage towards the wrong value of $\sigma^{2}$.

@@

The models are compared using Kullback-Leibler (KL) divergence from the distribution of good data to the posterior predictive distribution in order to assess the predictive capability of the models for the good part of the data. Specifically, for the $i^{th}$ group of the $k^{th}$ simulated data set $\by_{k}$ compute:
\begin{equation}
\label{kl}
KL^{(M)}_{ik} = \int \log \frac{f(\tilde y | \theta_{i}, \sigma^{2})}{f_{i}(\tilde y | M, \by_{k})}  f(\tilde y | \theta_{i}, \sigma^{2}) \ dy
\end{equation}
where $M$ indexes the fitting model, $f(\tilde y | \theta_{i}, \sigma^{2})$ is the pdf of a $N(\theta_{i}, \sigma^{2})$ at $\tilde y$ with $\sigma^{2} = 4$. For the Bayesian models ${f_{i}(\tilde y | M, \by_{k})} = \int f(\tilde y |\theta_{i}, \sigma_{i}^{2}) \pi(\theta_{i}, \sigma^{2}_{i} | M, \by_{k})$ where $ f(\tilde y |\theta_{i}, \sigma_{i}^{2})$ is  $N(\theta_{i}, \sigma_{i}^{2})$ pdf and  $\pi(\theta_{i}, \sigma^{2}_{i} | M, \by_{k})$ is the posterior for the $i^{th}$ group model parameters under model $M$ for the $k^{th}$ data set. $M$ denotes either the full normal theory model \eqref{fullsim2} or one of the two restricted likelihood versions discussed above, where conditioning is only on the summary statistics (not the complete $\by_{k}$). $M$ also specifies the hyperparameter values $a_{s}$ and $c_{s}$. For the classical robust fits, we set $f_{i}(\tilde y | M, \by_{k})$ to a $N(\hat\theta_{i}, \hat\sigma^{2}_{i})$ at  $\tilde y$ as a groupwise plug-in estimator for the predictive distribution. Thus the classical fits are done separately for each group - with no consideration of the hierarchical structure between the groups. The overall mean $\overline{KL}^{(M)}_{{\cdot}{\cdot}} = \frac{1}{90K} \sum_{k = 1}^{K} \sum_{i=1}^{90} KL^{(M)}_{ik}$ is used to compare the models. Sampling variation is summarized with $SE(\overline{KL}^{(M)}_{{\cdot} k}) = \sqrt{\frac{1}{K(K-1)}\sum_{k = 1}^{K} (\overline{KL}^{(M)}_{{\cdot} k} - \overline{KL}^{(M)}_{{\cdot}{\cdot}})^{2}}$ where $\overline{KL}^{(M)}_{{\cdot} k} = \frac{1}{90}\sum_{i = 1}^{90} KL^{(M)}_{ik}$.

Figure \ref{kl_sim} displays the results for each $a_{s} = 1.25,  5, 10$ and $c_{s} = 1/2, 1, 2$ over $K = 30$ data sets. The panels are for $c_{s} = 1/2, 1, 2$ with $a_{s} = 1.25,  5, 10$ on the abscissa. For each panel, the KL results for the classical robust linear model fits are the same since the priors have no effect on these fits. The normal theory model results are left out as they perform poorly. Overall, the results are strikingly in favor of the restricted likelihood methods for the range of hyper-parameter values studied. Undoubtably, the most precise and accurate prior studied is  $c_{s} = 1$ and $a_{s} = 10$ and this results in the lowest (and best) average KL for both the Tukey and Huber restricted likelihood versions. $c_{s} = 0.5$ performs the worst, but still better than the classical fits. There is evidence that performance starts to degrade as $a_{s} = 10$ as reflected in a larger average KL for $a_{s} = 10$. Here, the prior mean and precision are $2.22$ and $1.62$ and we suspect this is starting to put too much mass on $\sigma^{2}_{i}$ values much smaller than $\sigma^{2}=4$. For $c_{s} = 2$, the performance still improves from $a_{s} = 5$ to $a_{s} = 10$. Here the mean is $8.89$ and precision is only $0.1$; not yet large enough to degrade the performance due to an incorrect mean.

\begin{figure}[H]
\centering
\includegraphics[width = 6in, height = 4in]{kl_sim2_facet_scale.png}
\label{kl_sim}
\caption{summary of simulation}
\end{figure}

It is also interesting to look at the effects of $n_{i}$, $p_{i}$, and $m_{i}$. For this, the  $KL^{(M)}_{ik}$ are grouped by

%
%\begin{figure}[H]
%\centering
%\includegraphics[scale = .3]{kl_sim2_facet_as.png}
%\end{figure}
